\cleardoublepage
\phantomsection
\chapter*{Colophon}
\pagenumbering{gobble}

The author typeset the print version of this book in Dejavu and Source Code Pro
typefaces using the \TeX\ system. \LaTeX\ macros and {\hologo {XeTeX}}
extensions were used. Many other \TeX\ packages were also used in the
preparation of this book. The book uses two variants of the Dejavu typeface:
DejaVuSerif and DejaVuSans. It also uses SourceCodePro for monospaced text.

The picture used on the book cover is taken
from \href{https://commons.wikimedia.org/wiki/File:The_ladder_of_life_is_full_of_splinters.jpg}{Wikimedia
Commons}. The original photo is by Mykl Roventine and it is licensed under the
Creative Commons \href{https://creativecommons.org/licenses/by/2.0}{CC BY 2.0}
license.

The typefaces used in the cover design include Cabin, CabinCondensed, and
DejaVuSans.

Without the contributions of all the developers who work on free and open source
projects, it would not be possible to publish a book like this. Thank you to all
of these contributors for their hard work and dedication.

\section*{Reviews and Feedback}

The author would be very grateful for your reviews on Amazon and other websites.
When you tweet about the book, please include the hashtag \#essentialsofgo. Your
blog posts, videos, or anything else you create about this book will help to
spread the word. Your reviews and social media posts can help other people find
the book and learn about Go programming.


Please kindly send your feedback and suggestions to the following email address:
\href{mailto:baiju.m.mail@gmail.com}{baiju.m.mail@gmail.com}. You have the option
to create issues in the GitHub repository at:
\url{https://github.com/baijum/essential-go}. Additionally, you can propose changes by
creating pull requests.

Thank you for reviewing my book. Your feedback and suggestions are very valuable
and will help me improve the book. I will carefully consider your feedback and
make changes as needed. If you have any further questions or suggestions, please
do not hesitate to contact me.

\subsection*{Why should you review the book?}

\begin{enumerate}

\item This book will help many people learn Go programming. As a fellow Go community member, you are contributing to this effort.

\item Your feedback will influence the content of the book.

\item As you study the book with a critical mind, you are enhancing your Go knowledge. The learning is more intense than reading a book casually.

\item You will be required to refer to other works to verify the content in the text. This will help you to broaden your knowledge on the subject.

\item You are supporting a fellow Go community member. This will motivate me to contribute more.

\item Your name will be mentioned in the book at the beginning in the acknowledgments, just after the preface.

\item The book is a free or open source book licensed under the \href{http://creativecommons.org/licenses/by-sa/4.0/}{Creative Commons Attribution-ShareAlike 4.0 International} license.

\end{enumerate}

\subsection*{How to review the book?}

Reflect on these questions as you read the book.

\begin{enumerate}

\item How is the structure of the chapters in the book?
\item How are the sections under each chapter organized?
\item Is the book consistent throughout all the pages? (Consider consistency of organization, writing style, terminology usage, etc.)
\item Is there any essential point missing when explaining a particular topic?
\item Is the logical flow of introducing the concepts easy to follow?
\item Is the narrative clear, reasonable, and engaging?
\item Are the examples and exercises short and easy to understand?
\item Did you notice any sentence or paragraph that is difficult to understand?
\item Did you see an opportunity to add a side note that would be helpful to the reader?
\item Did you notice any awkward use of the English language? (You can ignore minor language issues.)
\item How can this book be made more interesting?

\end{enumerate}

When you provide feedback about a chapter, please specify the chapter title. If
your feedback is about a section, please include the section number and chapter
title. There is no need to mention the page number, as it may change.

If you are able to send a pull request to the Git repo with your changes, that
would be preferable to sending an email. If sending a pull request is not
possible, you can send an email as explained here.\\[40mm]

%% \pagebreak

\thispagestyle{empty} %% Remove header and footer.

\section*{About the Author}

Baiju Muthukadan is a software engineer from Kerala, India. He began his
software development career in 2002. Over the years, Baiju has contributed to
many free and open source software projects. In 2007, he authored his first book
on Python. Since 2013, he has been working on various Go projects. His Twitter
handle is \href{https://twitter.com/baijum}{@baijum}.
